---
title: "Untitled"
author: "Xuanqi Liang(U86768582)"
date: "2/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(ggthemes)
library(scales)
library(dplyr)
library(glmnet)
library(rpart) 
library(rpart.plot)
library(randomForest)
library(tidyverse)
library(caret)
library(ipred)# for fitting bagged decision trees
library(gbm)
library(boot)
library(hms)
theme_set(theme_bw())
```

## Load Dataset 

The "df" contains observations for the three airports. The other three are subset data tables for each airport respectively.

```{r}
df <- fread("/Users/liangxuanqi/Desktop/flights-master.csv")
summary(df)
```
```{r}
df$DAY_OF_MONTH <- as.factor(df$DAY_OF_MONTH)
df$YEAR <- as.factor(df$YEAR)
df$DAY_OF_WEEK <- as.factor(df$DAY_OF_WEEK)



# dropping columns for models
df[,MKT_CARRIER_AIRLINE_ID := NULL]
df[,MKT_CARRIER := NULL]
df[,TAIL_NUM := NULL]
df[,ORIGIN_CITY_NAME := NULL]
df[,ORIGIN_STATE_ABR := NULL]
df[,DEST_CITY_NAME := NULL]
df[,DEST_STATE_ABR := NULL]
df[,DEST_STATE_NM := NULL]
df[,DEP_TIME := NULL]
df[,CRS_ARR_TIME := NULL]
df[,ARR_TIME := NULL]
df[,ARR_DELAY := NULL]
df[,CANCELLED := NULL]
df[,DIVERTED := NULL]
df[,CRS_ELAPSED_TIME := NULL]
df[,ACTUAL_ELAPSED_TIME := NULL]
df[,AIR_TIME := NULL]
df[,DIV_AIRPORT_LANDINGS := NULL]
df[,DIV_REACHED_DEST := NULL]
df[,CODE := NULL]
df[,EARLY_AM := NULL]
df[,AM := NULL]
df[,PM := NULL]
df[,LATE_PM := NULL]

df.la <- df[ORIGIN == 'LAX', ]
df.bos <- df[ORIGIN == 'BOS', ]
df.atl <- df[ORIGIN == 'ATL', ]
```


```{r}
df.atl$CRS_DEP_TIME <- as.character(df.atl$CRS_DEP_TIME)
for( i in 1:nrow(df.atl)){
  x <- df.atl[i,]
  number_string = nchar(x$CRS_DEP_TIME)
  if(number_string == 3){
    df.atl[i,CRS_DEP_TIME := paste0("0", CRS_DEP_TIME)]
  }else if(number_string == 2){
    df.atl[i,CRS_DEP_TIME := paste0("0", CRS_DEP_TIME)]
    df.atl[i,CRS_DEP_TIME := paste0("0", CRS_DEP_TIME)]
  }else if(number_string == 1){
    df.atl[i,CRS_DEP_TIME := paste0("0", CRS_DEP_TIME)]
    df.atl[i,CRS_DEP_TIME := paste0("0", CRS_DEP_TIME)]
    df.atl[i,CRS_DEP_TIME := paste0("0", CRS_DEP_TIME)]
  }
}

d1 <- strptime(df.atl$CRS_DEP_TIME, format = "%H%M")
d2 <- format(d1, format = "%H:%M:%S")
df.atl[,CRS_DEP_TIME_Formatted := d2]
df.atl$CRS_DEP_TIME_Formatted <- as.factor(df.atl$CRS_DEP_TIME_Formatted)
df.atl[,CRS_DEP_HOUR := hour(as_hms(as.character(df.atl$CRS_DEP_TIME_Formatted)))]
```


```{r}
weather <- fread("/Users/liangxuanqi/Desktop/atlanta-weather.csv")
head(weather)
weather$day <- as.factor(weather$day)
weather$year <- as.factor(weather$year)

atlanta <- merge(df.atl, weather, all.x =TRUE, by.x = c('DAY_OF_MONTH','YEAR'), by.y = c('day','year'))
head(atlanta)
```
## Exploratory data analysis


```{r}
delayed <- df[DEP_DELAY > 0, .(number_delays = .N), by = ORIGIN]
not_delayed <- df[DEP_DELAY <= 0,.(number_on_time = .N), by= ORIGIN]
total <- merge(delayed, not_delayed, by='ORIGIN')
total$ORIGIN <- as.factor(total$ORIGIN)
dat_long <- total %>%
  gather("Stat", "Value", -ORIGIN)
setDT(dat_long)
ggplot(dat_long, aes(x = ORIGIN, y = Value, fill = Stat)) +
  geom_col(position = "dodge") + ylab('Number of flights')  + ggtitle("Delay vs. On-time flights")
```
Across all three locations, the majority of flights are on time. The trick is being able to predict which flights will be delayed.
```{r}
flights_by_year <-df[,.(Number_flights = .N), by =.(ORIGIN, YEAR)]
flights_by_year$YEAR <- as.factor(flights_by_year$YEAR)
ggplot(flights_by_year, aes(x = ORIGIN, y = Number_flights, fill =YEAR)) +
  geom_col(position = "dodge") + ylab('Number of flights') + ggtitle("Number of flights - 2019 vs. 2020")
```

```{r}
weekly_delay <-df[,.(mean_delay = mean(DEP_DELAY)), by =.(DAY_OF_WEEK = df$DAY_OF_WEEK, ORIGIN = df$ORIGIN)]
ggplot(weekly_delay,aes(x = DAY_OF_WEEK, y = mean_delay, color=ORIGIN)) + geom_point() + geom_line()  + ylab('Average Delay') + xlab('Day of the week') + ggtitle("Averga delay in days of the week")
```

Delays tend to increase later in the week. This could be because airports are busier on the weekend. We will likely want to keep Day of Week as a predictor in our models.

```{r}
day_month_delay <-  df[,.(mean_delay = mean(DEP_DELAY)), by =.(DAY_OF_MONTH = DAY_OF_MONTH, ORIGIN = ORIGIN)]
ggplot(day_month_delay,aes(x = DAY_OF_MONTH, y = mean_delay, color = ORIGIN)) + geom_point() + geom_line() + ylab('Average Delay') + xlab('Day of the month') + ggtitle("Average delay in days of the month")
```
Boston has some outliers skewing the avg delay. Otherwise, all three locations share similar trends across the month.

```{r}
distance_delay_weekly <- df.atl[,.(delay = mean(DEP_DELAY), dist = mean(DISTANCE)), by ='DAY_OF_WEEK']
ggplot(distance_delay_weekly,aes(x = dist, y = delay)) + geom_point() + geom_smooth(method = "lm") + ggtitle("Los Angeles - Average Departure and distance by day of the week") + ylab("Average Departure Delay") + xlab("Average Flight Distance")
distance_delay_airline <- df.atl[,.( delay = mean(DEP_DELAY), dist = mean(DISTANCE)), by ='AIRLINE']
ggplot(distance_delay_airline,aes(x = dist, y = delay)) + geom_point() + geom_smooth(method = "lm") + ggtitle("Los Angeles - Average Departure and flight distance by airline") + ylab("Average Departure Delay") + xlab("Average Flight Distance")
```

## Models for each airport

In this section, we present many of our models for each airport location and highlight which perform best (some models' code may be excluded in order to keep the length of the document down!)

### Los Angeles Airport

```{r echo=FALSE}
set.seed(123)
smp_size <- floor(.80 * nrow(atlanta))
train_index <- sample(nrow(atlanta), smp_size) # assign 80% of the data to train
df.test <- atlanta[-train_index,] ##not the one in train_index
df.train <-atlanta[train_index,] 
```

#### Naive Regression

Started doing the Naive Regression (Baseline) to have an MSE to compare to. 

```{r}
set.seed(123)
y.test.b <- df.test$DEP_DELAY
df_atl_y <- mean(df.train$DEP_DELAY)
mse_baseline <- mean((y.test.b - df_atl_y)^2)
rmse_baseline <- sqrt(mse_baseline)
mse_baseline
rmse_baseline 
```

#### Linear Regression

##### Validation set approach

Fit the model and make predictions using the train & test data sets. 

```{r}
y.train.lm <- df.train$DEP_DELAY
y.test.lm <- df.test$DEP_DELAY
f1 <- as.formula(DEP_DELAY ~ DAY_OF_MONTH + YEAR + DAY_OF_WEEK + AIRLINE + CRS_DEP_HOUR + humidity + precipMM + pressure + tempC + visibility + windspeedKmph)
fit.lm1 <- lm(f1,df.train) 
#Let's compute an MSE on the training data
yhat.train.lm1 <- predict(fit.lm1)
mse.train.lm1 <- mean((y.train.lm - yhat.train.lm1)^2)
mse.train.lm1
#Let's compute an MSE on the test data
yhat.test.lm1 <- predict(fit.lm1, df.test)
mse.test.lm1 <- mean((y.test.lm - yhat.test.lm1)^2)
mse.test.lm1
rmse_lm1 <- sqrt(mse.test.lm1)
rmse_lm1
```

##### K-folds cross validation

Considered the LOOCV method but it was too much computational cost. We decided to do the K-folds approach. 

```{r}
train.control <- trainControl(method = "cv", number = 10)
#train.control <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
model <- train(f1, data =atlanta, method = "lm", trControl = train.control) # Train the model
print(model) # Summarize the results
```

#### Ridge Regression

Ridge regression shrinks coefficients towards zero. We used glmnet command to shrink coefficients towards zero. We passed the predictors matrix as parameters and used alpha=0 to invoke ridge regression. 

```{r}
dd <- copy(atlanta)
dd[, test:=0] #Adds a new column with value 0
dd[sample(nrow(dd),5000), test:=1] #Take 5k random rows and assign it to test
dd.test <- dd[test==1]
dd.train <- dd[test==0] #Around 33K for training
#assign our response variable (target)
y.train.r <- dd.train$DEP_DELAY
y.test.r <- dd.test$DEP_DELAY
x1.train <- model.matrix(f1, dd.train)[,-1]
x1.test <- model.matrix(f1,dd.test)[,-1]
fit.ridge <- cv.glmnet(x1.train, y.train.r, alpha = 0, nfolds = 10)
##Test the MSE in training data
yhat.train.ridge <- predict(fit.ridge, x1.train, s = fit.ridge$lambda.min)
mse.train.ridge <- mean((y.train.r - yhat.train.ridge)^2)
mse.train.ridge
##Test the MSE test
yhat.test.ridge <- predict(fit.ridge, x1.test, alpha = 0, s = fit.ridge$lambda.min)
mse.test.ridge <-  mean((y.test.r - yhat.test.ridge)^2)
mse.test.ridge 
rsme_test_ridge <- sqrt(mse.test.ridge)
rsme_test_ridge 
```

```{r}
optimal_lambda <- fit.ridge$lambda.min
coef(fit.ridge, s = optimal_lambda)
```

The optimal lambda is `r optimal_lambda`. 


#### Lasso Regression

The lasso is like ridge regression – but instead of shrinking coefficients towards zero,it tries to set as many as it can to zero.

```{r}
dd.lasso <- copy(atlanta)
dd.lasso[, test:=0] 
dd.lasso[sample(nrow(dd.lasso),5000), test:=1]
dd.test.l <- dd.lasso[test==1]
dd.train.l <- dd.lasso[test==0] #Around 33K for training
y.train.lasso <- dd.train.l$DEP_DELAY
y.test.lasso <- dd.test.l$DEP_DELAY
x2.train <- model.matrix(f1, dd.train.l)[,-1]
x2.test <- model.matrix(f1,dd.test.l)[,-1]
fit.lasso <- cv.glmnet(x2.train, y.train.lasso, alpha = 1, nfolds = 10)
##Test the MSE in training data
yhat.train.lasso <- predict(fit.lasso, x2.train, s = fit.lasso$lambda.min)
mse.train.lasso <- mean((y.train.lasso - yhat.train.lasso)^2)
mse.train.lasso
##Test the MSE test
yhat.test.lasso <- predict(fit.lasso, x2.test, alpha = 0, s = fit.lasso$lambda.min)
mse.test.lasso <-  mean((y.test.lasso - yhat.test.ridge)^2)
mse.test.lasso 
rmse_lasso <- sqrt(mse.test.lasso)
rmse_lasso
```
#### Decision Tree Model

```{r}
smp_size <- floor(.75 * nrow(atlanta))
train_index <- sample(nrow(atlanta), smp_size) 
df.test.dt <- atlanta[-train_index,] ##not the one in train_index
df.train.dt <-atlanta[train_index,] 
y.train.dt <- df.train.dt$DEP_DELAY
y.test.dt <- df.test.dt$DEP_DELAY
f5 <- as.formula(DEP_DELAY ~ DAY_OF_MONTH + YEAR + CRS_DEP_HOUR  + CRS_ARR_TIME 
                 + CRS_ELAPSED_TIME + humidity + precipMM)
#grow tree
fit.tree <- rpart(f1, data = df.train.dt,control = rpart.control(cp = 0.001), method = "anova")
printcp(fit.tree) # display the results
plotcp(fit.tree) # visualize cross-validation results
```
A good choice of cp for pruning is often the leftmost value for which the mean lies below the horizontal line.
```{r}
# plot tree
rpart.plot(fit.tree, type = 1)
```

```{r}
# prune the tree
optimal_cp <- fit.tree$cptable[which.min(fit.tree$cptable[,"xerror"]),"CP"]
pfit<- prune(fit.tree, cp=optimal_cp) # from cptable   
rpart.plot(pfit)
summary(pfit)
yhat.train.tree <- predict(pfit, df.train.dt) 
mse.train.tree <- mean((y.train.dt - yhat.train.tree) ^ 2)
mse.train.tree
yhat.test.tree <- predict(pfit, df.test.dt) 
mse.test.tree <- mean((y.test.dt - yhat.test.tree) ^ 2)
mse.test.tree
rmse_tree_test <- sqrt(mse.test.tree)
rmse_tree_test 
```
