---
title: "Predicting flight delays - Los Angeles airport"
author: "Maraline Torres"
date: "2/27/2021"
output: html_document
---
## Introduction

Flying doesn't always go smoothly. Many people have had some horror stories with weird delays but other haven't experienced this (Lucky them!). This brings us to the following question:  Wouldn't it be nice to know how much your flight will probably be delayed?

That what this project will attempt to do, specifically for all airlines in Los Angeles Airport. We gathered data for flights in March 2019 and 2020. This will be a regression problem where we will try to predict delay time in number of minutes. 

## Getting the data

We gathered our flight data from the US Department of Transportation. We chose the following features:

* DAY_OF_MONTH
* YEAR
* DAY_OF_WEEK
* DEP_DELAY
* CRS_ARR_TIME
* CRS_ELAPSED_TIME
* CRS_DEP_HOUR 
* AIRLINE
* humidity
* precipMM
* pressure
* tempC
* visibility
* windspeedKmph 

The weather variables were retrieved using the World Weather API. 

## Load libraries
```{r}
library(data.table)
library(ggplot2)
library(ggthemes)
library(scales)
library(dplyr)
library(glmnet)
library(rpart) 
library(rpart.plot)
library(randomForest)
library(tidyverse)
library(caret)
library(ipred)# for fitting bagged decision trees
library(gbm)
library(boot)

theme_set(theme_bw())
```

## Load Dataset

```{r}
df.la <- fread("la_flights.csv")
summary(df)
```


## Exploratory data analysis



## Start fitting models


### Splitting the datasets
```{r}
set.seed(123)
smp_size <- floor(.80 * nrow(df.la))
train_index <- sample(nrow(df.la), smp_size) # assign 80% of the data to train
df.test <- df.la[-train_index,] ##not the one in train_index
df.train <-df.la[train_index,] 
```

### Naive Regression

Started doing the Naive Regression (Baseline) to have an MSE to compare to. 

```{r}
y.test.b <- df.test$DEP_DELAY
df_la_y <- mean(df.la$DEP_DELAY)

mse_baseline <- mean((y.test.b - df_la_y)^2)
rmse_baseline <- sqrt(mse_baseline)
rmse_baseline 
```

Our goal is to apply more advanced ML methods  to obtain a lower RMSE.

### Linear Regression

#### Validation set approach

Fit the model and make predictions using the two data sets created earlier. 

```{r}
y.train <- df.train$DEP_DELAY
y.test <- df.test$DEP_DELAY

f1 <- as.formula(DEP_DELAY ~ DAY_OF_MONTH + YEAR + DAY_OF_WEEK + AIRLINE 
                 + CRS_DEP_HOUR  + CRS_ARR_TIME + humidity + precipMM + pressure 
                 + tempC + visibility + windspeedKmph)

fit.lm1 <- lm(f1,df.train) 

#Let's compute an MSE on the training data
yhat.train.lm1 <- predict(fit.lm1)
mse.train.lm1 <- mean((y.train - yhat.train.lm1)^2)
mse.train.lm1

#Let's compute an MSE on the test data
yhat.test.lm1 <- predict(fit.lm1, df.test)
mse.test.lm1 <- mean((y.test - yhat.test.lm1)^2)
mse.test.lm1

rmse_lm1 <- sqrt(mse.test.lm1)
rmse_lm1

```


Create a new formula for the variables that had significant value. 

```{r}
summary(fit.lm1)

f2 <- as.formula(DEP_DELAY ~ DAY_OF_MONTH + YEAR + DAY_OF_WEEK + AIRLINE 
                 + CRS_DEP_HOUR  + CRS_ARR_TIME + humidity)

fit.lm2 <- lm(f2,df.train) 


#Let's compute an MSE on the training data
yhat.train.lm2 <- predict(fit.lm1)
mse.train.lm2 <- mean((y.train - yhat.train.lm2)^2)
mse.train.lm2

#Let's compute an MSE on the test data
yhat.test.lm2 <- predict(fit.lm2, df.test)
mse.test.lm2 <- mean((y.test - yhat.test.lm2)^2)
mse.test.lm2

rmse_lm1 <- sqrt(mse.test.lm2)
rmse_lm1
```

#### K-folds cross validation

Considered making a

```{r}
train.control <- trainControl(method = "cv", number = 10)
#train.control <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
model <- train(f2, data = df.la, method = "lm", trControl = train.control) # Train the model
print(model) # Summarize the results
```

We present a similar RSME as the other formulas for linear regression. We decided that we need a more flexible model to explain our observations. 


### Ridge Regression

Ridge regression shrinks coefficients towards zero. It uses glmnet commando to shrink coefficients towards zero. We passed the predictors matrix as parameters and used alpha=0 to invoke ridge regression. 

```{r}
dd <- copy(df.la)
dd[, test:=0] #Adds a new column with value 0
dd[sample(nrow(dd),5000), test:=1] #Take 5k random rows and assign it to test
dd.test <- dd[test==1]
dd.train <- dd[test==0] #Around 33K for training

f3 <- as.formula(DEP_DELAY ~ DAY_OF_MONTH + YEAR + DAY_OF_WEEK + AIRLINE 
                 + CRS_DEP_HOUR  + CRS_ARR_TIME + humidity)

#assign our response variable (target)
y.train <- dd.train$DEP_DELAY
y.test <- dd.test$DEP_DELAY

x3.train <- model.matrix(f3, dd.train)[,-1]
x3.test <- model.matrix(f3,dd.test)[,-1]

fit.ridge <- cv.glmnet(x3.train, y.train, alpha = 0, nfolds = 10)

##Test the MSE in training data
yhat.train.ridge <- predict(fit.ridge, x5.train, s = fit.ridge$lambda.min)
mse.train.ridge <- mean((y.train - yhat.train.ridge)^2)
mse.train.ridge

##Test the MSE test
yhat.test.ridge <- predict(fit.ridge, x3.test, alpha = 0, s = fit.ridge$lambda.min)
mse.test.ridge <-  mean((y.test - yhat.test.ridge)^2)
mse.test.ridge 

rsme_test_ridge <- sqrt(mse.test.ridge)
rsme_test_ridge 

```

### Lasso Regression

The lasso is like ridge regression – but instead of shrinking coefficients towards zero,it tries to set as many as it can to zero.

```{r}
dd <- copy(df.la)
dd[, test:=0] 
dd[sample(nrow(dd),5000), test:=1]
dd.test <- dd[test==1]
dd.train <- dd[test==0] #Around 33K for training

y.train <- dd.train$DEP_DELAY
y.test <- dd.test$DEP_DELAY

f4 <- as.formula(DEP_DELAY ~ DAY_OF_MONTH + YEAR + DAY_OF_WEEK + AIRLINE 
                 + CRS_DEP_HOUR  + CRS_ARR_TIME + humidity)

x4.train <- model.matrix(f4, dd.train)[,-1]
x4.test <- model.matrix(f4,dd.test)[,-1]

fit.lasso <- cv.glmnet(x4.train, y.train, alpha = 1, nfolds = 10)

##Test the MSE in training data
yhat.train.lasso <- predict(fit.lasso, x4.train, s = fit.lasso$lambda.min)
mse.train.lasso <- mean((y.train - yhat.train.lasso)^2)
mse.train.lasso

##Test the MSE test
yhat.test.lasso <- predict(fit.lasso, x4.test, alpha = 0, s = fit.lasso$lambda.min)
mse.test.lasso <-  mean((y.test - yhat.test.ridge)^2)
mse.test.lasso 

rsme_lasso <- sqrt(mse.test.lasso)

```


### DECISION TREE MODEL

```{r}
set.seed(123)
smp_size <- floor(.80 * nrow(df.la))
train_index <- sample(nrow(df.la), smp_size) 
df.test <- df.la[-train_index,] ##not the one in train_index
df.train <-df.la[train_index,] 

y.train <- df.train$DEP_DELAY
y.test <- df.test$DEP_DELAY

f5 <- as.formula(DEP_DELAY ~ DAY_OF_MONTH + YEAR + CRS_DEP_HOUR  + CRS_ARR_TIME 
                 + CRS_ELAPSED_TIME + humidity + precipMM)

#grow tree
fit.tree <- rpart(f5, data = df.train,control = rpart.control(cp = 0.001), method = "anova")
printcp(fit.tree) # display the results
plotcp(fit.tree) # visualize cross-validation results
#A good choice of cp for pruning is often the leftmost value 
#for which the mean lies below the horizontal line." PS library(rpart)
summary(fit.tree) # detailed summary of splits

# create additional plots
par(mfrow=c(1,2)) # two plots on one page
rsq.rpart(fit.tree) # visualize cross-validation results  

# plot tree
rpart.plot(fit.tree)
rpart.plot(fit.tree, type = 1)

# prune the tree
pfit<- prune(fit.tree, cp=0.0028032) # from cptable   
rpart.plot(pfit)
summary(pfit)

yhat.train.tree <- predict(pfit, df.train) 
mse.train.tree <- mean((y.train - yhat.train.tree) ^ 2)
mse.train.tree

yhat.test.tree <- predict(pfit, df.test) 
mse.test.tree <- mean((y.test - yhat.test.tree) ^ 2)
mse.test.tree

rmse_tree_test <- sqrt(mse.test.tree)
rmse_tree_test 


```


### Random Forrest 

```{r}
dd <- copy(df.la)
dd[, test:=0]
dd[sample(nrow(dd), 5000), test:=1] # take 5K random rows and stick them in the test
dd.test <- dd[test==1]
dd.train <- dd[test==0]

dd.train.sample.size <- 25000
dd.train.sample <- dd.train[sample(nrow(dd.train), dd.train.sample.size)]

y.train <- dd.train$DEP_DELAY
y.train.sample <- dd.train.sample$DEP_DELAY
y.test <- dd.test$DEP_DELAY

f6 <- as.formula(DEP_DELAY ~ DAY_OF_MONTH + AIRLINE + CRS_DEP_HOUR  + CRS_ARR_TIME 
                 + CRS_ELAPSED_TIME)
fit.rndfor <- randomForest(f6, dd.train,ntree=500, do.trace=F)

print(fit.rndfor) #View results
importance(fit.rndfor)
#We can check which variables are most predictive using a variable importance plot
varImpPlot(fit.rndfor) #Arrival time, CRS Elapsed time, Depature hour, Airline and Day of Month


yhat.train.rndfor <- predict(fit.rndfor, dd.train.sample) 
mse.train.rndfor <- mean((y.train.sample - yhat.train.rndfor) ^ 2)
mse.train.rndfor

yhat.test.rndfor <- predict(fit.rndfor, dd.test) 
mse.test.rndfor <- mean((y.test - yhat.test.rndfor) ^ 2)
mse.test.rndfor

rmse.test.rndfor <- sqrt(mse.test.rndfor)
rmse.test.rndfor

```

